---
title: "Data Visualization with Pie Charts in R"
author: "julius"
format: docx
---

# Introduction

Pie charts are a popular tool in data visualization used to represent the proportions of different categories within a whole. They are simple, visually appealing, and effective for displaying relative sizes of data points.

## Importance of Pie Charts

* Representing Proportions

They provide a clear visual representation of how different categories contribute to the whole.

* Simplifying Data

They make it easy to understand the composition of a dataset at a glance.

* Highlighting Differences

They effectively highlight differences in proportions among categories.

* Engaging Visuals

Their circular shape and colorful segments make them visually engaging and easy to interpret.

## Use Cases

-   Market Share Analysis: Pie charts are commonly used to represent market share, showing the percentage of total sales or revenue each company contributes.

-   Budget Allocation: Pie charts effectively communicate how a budget is distributed among different expense categories.

-   Survey Responses: When presenting survey results, pie charts can illustrate the distribution of responses across various answer choices.

## Best Practices for Pie Charts

-   Limit number of slices

Too many slices can make the chart cluttered and hard to read. ideally keep it to five or fewer categories

-   Use Distinct Colors Ensure each slice is easily distinguishable by useing distinct colors

-   Label Clearly Each slice should be clealry labeled with its category and percentage

-   Maintain Proportionality Ensures the size of the slices are accurately proportional to the values they represent

-   Consider Alternatives For datasets with many categories consider using other charts each bar chats

## Creating Pie Charts in R

To create pie charts in R, we will use the ggplot2 package in r. Let's start with a basic pie chart


```{r,message=FALSE,warning=FALSE}
library(tidyverse)

# Create a sample dataset
data <- data.frame(
  Category = c("A", "B", "C", "D"),
  Value = c(30, 20, 25, 25)
)

# Create a basic pie chart
ggplot(data, aes(x = "", y = Value, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  labs(title = "Basic Pie Chart") +
  theme_void()

```

### Pie Chart with Labels

Adding labels to a pie chart provides additional information about the proportions of each category.

```{r}
# Create a pie chart with labels
ggplot(data, aes(x = "", y = Value, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = Value), 
            position = position_stack(vjust = 0.5)) +
  labs(title = "Pie Chart with Labels") +
  theme_void()

```

### Colored Pie Chart

Using different colors for different categories enhances the visual distinction between segments.

```{r}
# Create a colored pie chart
ggplot(data, aes(x = "", y = Value, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  scale_fill_manual(values = c("red", "blue", "green", "purple")) +
  labs(title = "Colored Pie Chart") +
  theme_void()

```

### Donut Chart

A donut chart is a variation of a pie chart with a blank center, which can make it easier to compare the sizes of segments.

```{r}
# Create a donut chart
ggplot(data, aes(x = 2, y = Value, fill = Category)) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  labs(title = "Donut Chart") +
  theme_void() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

```

```{r}
# library(plotly)
# ds <- data.frame(labels = c("A", "B", "C"), values = c(10, 40, 60))
# plot_ly(ds, labels = ~labels, values = ~values) %>%
#   add_pie(hole = 0.2) %>%
#   layout(title = "Pie Chart using Plotly")
```

### Pie Chart with Percentage Labels

Displaying percentages on a pie chart helps to clearly communicate the proportion of each segment.

```{r}
# Create a pie chart with percentage labels
data$Percentage <- round(data$Value / sum(data$Value) * 100)

ggplot(data, aes(x = "", y = Value, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percentage, "%")), 
            position = position_stack(vjust = 0.5)) +
  labs(title = "Pie Chart with Percentage Labels") +
  theme_void()

```
